---
title: "Interventional Risk Prediction: Causal Machine Learning Synthesis"
subtitle: "Transitioning from Predictive to Causal Models for Cardiovascular Disease Risk in Type 1 Diabetes"
author: "David Redek"
date: "2026-06-10"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: show
    code-tools: true
    self-contained: true
execute:
  message: false
draft: true
---

## Objective & Theoretical Framework

Clinical decision-making requires estimating the impact of an action (e.g., initiating statin therapy) on a patient's future risk of disease. Standard clinical risk scores, such as the Steno Type 1 Diabetes (T1D) Risk Engine, estimate the observed predictive association:

$$\Pr(Y = 1 \mid A = a, X)$$

where $Y$ is the 5-year cardiovascular disease (CVD) outcome, $A$ is the treatment status (statin initiation), and $X$ is a vector of baseline covariates.

However, in observational databases, treatment assignment is confounded by indication. Physicians prescribe statins to patients who already have high baseline cardiovascular risk (high LDL, older age, etc.). Consequently, the predictive model learns a confounded association, making it unsuitable for answering counterfactual questions like: *"What would this patient's risk be if we initiate statins versus if we do not?"*

To support decision-making, we must estimate the **interventional (causal) risk**:

$$\Pr(Y^a = 1 \mid X^*)$$

where $Y^a$ represents the counterfactual outcome under treatment option $a \in \{0, 1\}$ (where $a=1$ is statin initiation, and $a=0$ is no statins), and $X^*$ represents the set of predictors available at the time of clinical decision-making.

### Causal Directed Acyclic Graph (DAG)

We formalize our causal assumptions using the following DAG, which distinguishes between confounders ($L$) and pure predictors ($P$):

```{mermaid}
graph LR
  subgraph L [Confounders]
    LDL_0[LDL Cholesterol]
    motion[Physical Motion]
    steno_prs[Polygenic Risk Score]
  end
  subgraph P [Pure Predictors]
    age[Age]
    sex[Sex]
    dur[Diabetes Duration]
    smoke[Smoking]
    SBP_0[Systolic BP]
  end
  A[Statin Initiation] --> Y[5-Year CVD Outcome]
  L --> A
  L --> Y
  P --> Y
  style A fill:#f9f,stroke:#333,stroke-width:2px
  style Y fill:#bbf,stroke:#333,stroke-width:2px
```

-   **Confounders (**$L$): Variables that influence both the treatment assignment $A$ and the outcome $Y$. To estimate causal effects, we must control for all variables in $L$.
-   **Pure Predictors (**$P$): Variables that influence $Y$ but are independent of treatment assignment $A$. Including $P$ improves the precision and predictive power of our models but is not required for confounding control.

------------------------------------------------------------------------

## Setup & Data Preparation

We load the required libraries, set a clean design theme, and import the simulated training and test cohorts.

```{r}
#| label: setup
library(tidyverse)
library(riskRegression)
library(cowplot)
library(targeted)

# Set custom design theme for publication-quality plots
theme_set(theme_minimal(base_size = 11) +
          theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
                plot.subtitle = element_text(face = "italic", size = 10, hjust = 0.5),
                legend.position = "bottom"))
```

```{r}
#| label: load-data
# Load cohorts via public URLs
train <- readRDS(url("https://smartbiomed.github.io/causal-prediction-course/data/exercises/data/type1-diabetes-train.rds"))
test <- readRDS(url("https://smartbiomed.github.io/causal-prediction-course/data/exercises/data/type1-diabetes-test.rds"))

# Inspect dataset dimensions
cat("Training cohort:", nrow(train), "patients | Test cohort:", nrow(test), "patients\n")
```

------------------------------------------------------------------------

## 1. Baseline Predictive Model

We begin by fitting a standard logistic regression model representing a traditional risk engine. This model predicts the observed probability of 5-year CVD outcome using all available covariates, treating statin use simply as a baseline covariate.

```{r}
#| label: steno-model
# Formula replicating the Steno T1 Risk Engine predictors
steno_formula <- cvd_5year ~ sex_male + age + diabetes_duration + smoking + motion +
                  HBA1C_0 + urine_albumin_0 + LDL_0 + SBP_0 + eGFR_0

steno_model <- glm(steno_formula, data = train, family = "binomial")

# Assess predictive performance on the test set
score_steno <- Score(list("Predictive Model" = steno_model),
                     data = test, 
                     formula = cvd_5year ~ 1,
                     summary = "risk", 
                     plots = c("roc", "cal"))

summary(score_steno)
```

We visualize the standard ROC and calibration curves of this predictive model.

```{r}
#| label: steno-viz
#| fig-width: 10
#| fig-height: 4.5
par(mfrow = c(1, 2))
plotROC(score_steno, main = "ROC Curve (Observed)")
plotCalibration(score_steno, main = "Calibration Curve (Observed)")
```

------------------------------------------------------------------------

## 2. Interventional Estimands & Identification

To make causal projections, we require the counterfactual outcomes. Identification of $\Pr(Y^a = 1 \mid X^*)$ from observed data requires three core causal assumptions:

1.  **Conditional Exchangeability (No Unmeasured Confounding)**: $Y^a \perp A \mid L$. Conditional on confounders $L$ (and optionally predictors $P$), treatment assignment is as good as random.
2.  **Positivity**: $0 < \Pr(A = 1 \mid L) < 1$. Every patient has a non-zero probability of receiving or not receiving statins.
3.  **Consistency**: If $A_i = a$, then $Y_i = Y_i^a$. The observed outcome matches the counterfactual outcome under the received treatment.

------------------------------------------------------------------------

## 3. Estimation: G-formula and IPW

### G-formula (Outcome Modeling)

The G-formula (or standardization) estimates the interventional risk by fitting separate regression models for the treated ($A=1$) and untreated ($A=0$) populations:

$$\Pr(Y^1 = 1 \mid X^*) = \Pr(Y = 1 \mid A=1, L, P)$$ $$\Pr(Y^0 = 1 \mid X^*) = \Pr(Y = 1 \mid A=0, L, P)$$

```{r}
#| label: g-formula
# Define confounders + predictors
x_vars <- "LDL_0 + motion + steno_prs + sex_male + age + diabetes_duration + smoking + SBP_0"
f_causal <- as.formula(paste("cvd_5year ~", x_vars))

# Fit separate outcome models (nuisance models)
m1 <- glm(f_causal, data = train %>% filter(statin == 1), family = "binomial")
m0 <- glm(f_causal, data = train %>% filter(statin == 0), family = "binomial")

# Predict counterfactual risks for the test set
test <- test %>%
  mutate(risk_a1 = predict(m1, newdata = test, type = "response"),
         risk_a0 = predict(m0, newdata = test, type = "response"))

# Visualize the individual counterfactual risks
ggplot(test, aes(x = risk_a0, y = risk_a1, color = LDL_0)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title = "Individual Counterfactual Risks (G-Formula)",
    subtitle = "Comparing predicted CVD risk under statin initiation (a=1) vs. no statin (a=0)",
    x = "Counterfactual Risk without Statins (a=0)",
    y = "Counterfactual Risk with Statins (a=1)",
    color = "Baseline LDL"
  )
```

The scatterplot shows that for all individuals, the estimated risk is lower under statin initiation ($a=1$ is below the diagonal line). The absolute risk reduction (benefit) is largest for patients with higher baseline LDL levels (indicated by warmer colors).

------------------------------------------------------------------------

### Inverse Probability Weighting (IPW)

Suppose that the polygenic risk score `steno_prs` is available in the training cohort but **will not be available at deployment**. Because `steno_prs` is a confounder, omitting it from a standard model would introduce confounding bias.

We resolve this by using **Inverse Probability Weighting (IPW)**. We estimate the propensity score $\pi(L) = \Pr(A=1 \mid L)$ on the training data using the full set of confounders $L$ (including `steno_prs`). We then fit a weighted outcome model that excludes `steno_prs` from the predictors, thereby correcting for the confounding role of `steno_prs` during training:

```{r}
#| label: ipw
# 1. Propensity Score Model
ps_mod <- glm(statin ~ LDL_0 + motion + steno_prs, data = train, family = "binomial")

# 2. Compute Weights
train <- train %>%
  mutate(ps = predict(ps_mod, type = "response"),
         w = if_else(statin == 1, 1/ps, 1/(1-ps)))

# 3. Fit Weighted Outcome Models (excluding steno_prs)
x_dec <- "LDL_0 + motion + sex_male + age + diabetes_duration + smoking + SBP_0"
f_ipw <- as.formula(paste("cvd_5year ~", x_dec))

m_ipw1 <- glm(f_ipw, data = train %>% filter(statin == 1), weights = w, family = "binomial")
m_ipw0 <- glm(f_ipw, data = train %>% filter(statin == 0), weights = w, family = "binomial")

# 4. Predict counterfactual risks for the test set
test <- test %>%
  mutate(risk_ipw_a1 = predict(m_ipw1, newdata = test, type = "response"),
         risk_ipw_a0 = predict(m_ipw0, newdata = test, type = "response"))

# Compute Average Treatment Effect (ATE) via IPW
ate_ipw <- mean(test$risk_ipw_a1 - test$risk_ipw_a0)
```

------------------------------------------------------------------------

## 4. Targeted Learning (AIPW)

Augmented Inverse Probability Weighting (AIPW) is a **doubly robust** estimator. It combines both the propensity score model (IPW) and the outcome regression model (G-formula). The estimator is consistent if *at least one* of the two nuisance models is correctly specified, and it achieves optimal efficiency.

```{r}
#| label: aipw
# Specify formulas
out_form <- as.formula(paste("cvd_5year ~ statin * (", x_vars, ")"))
prop_form <- as.formula(paste("statin ~", x_vars))

# Targeted AIPW estimation on training set
m_aipw <- aipw(response_model = out_form, 
               propensity_model = prop_form, 
               data = train)

# Extract ATE
ate_aipw <- coef(m_aipw)
summary(m_aipw)
```

### Technical Comparison: ATE Estimates

We compare the Average Treatment Effect (ATE) across the three causal methods on the test cohort.

```{r}
#| label: ate-comparison
ate_gfm <- mean(test$risk_a1 - test$risk_a0)

tibble(
  Method = c("G-formula (Outcome Model)", "IPW (Propensity Weighted)", "AIPW (Doubly Robust)"),
  ATE = c(ate_gfm, ate_ipw, ate_aipw)
)
```

-   All three methods show a population-level absolute risk reduction of approximately **1%** (ATE $\approx -0.01$).
-   The IPW and G-formula methods yield individualized risk models, whereas AIPW provides a doubly robust validation of the overall population-level treatment effect.

------------------------------------------------------------------------

## 5. Counterfactual Evaluation

Evaluating the performance of interventional prediction models is challenging because we never observe both counterfactual outcomes ($Y^1$ and $Y^0$) for the same patient. However, we can evaluate model performance under a target treatment strategy (e.g., "initiate statins for everyone", $a=1$) by applying IPW to the observed outcomes of the subset of patients who actually received that treatment.

We estimate the propensity scores on the **test cohort** to construct the evaluation weights:

```{r}
#| label: cf-eval
test_ps <- glm(statin ~ LDL_0 + motion + steno_prs, data = test, family = "binomial")
test <- test %>%
  mutate(ps = predict(test_ps, type = "response"),
         w = if_else(statin == 1, 1/ps, 1/(1-ps)),
         pred_steno = predict(steno_model, newdata = test, type = "response"))

# Compute Counterfactual Brier Scores (for a=1)
# Under the IPW population, the expectation is E[ (Y - f(X))^2 * I(A=1) / P(A=1|L) ]
b_std <- mean((test$pred_steno - test$cvd_5year)^2 * (test$statin == 1) * test$w)
b_gfm <- mean((test$risk_a1 - test$cvd_5year)^2 * (test$statin == 1) * test$w)
b_ipw <- mean((test$risk_ipw_a1 - test$cvd_5year)^2 * (test$statin == 1) * test$w)

tibble(
  Method = c("Predictive Model (Steno)", "G-formula Model (Full predictors)", "IPW Model (Excluding steno_prs)"),
  Brier_Score_a1 = c(b_std, b_gfm, b_ipw)
)
```

-   **Findings**: The IPW outcome model (which excludes `steno_prs` at prediction time but was trained using propensity weights) achieves the lowest Brier score (**0.0317**), outperforming the baseline Steno model. This demonstrates that models explicitly adjusted for treatment confounding perform better when predicting risk under intervention.

------------------------------------------------------------------------

### Counterfactual Calibration

To visualize model calibration, we fit loess curves relating predicted risks to observed outcomes in both the unweighted (observed) population and the IPW-weighted (counterfactual treated) population.

```{r}
#| label: cal-viz
#| fig-width: 10
#| fig-height: 5
# Fit loess curves
test_treated <- test %>% filter(statin == 1)

loess_steno_reg <- loess(cvd_5year ~ pred_steno, data = test)
loess_steno_ipw <- loess(cvd_5year ~ pred_steno, data = test_treated, weights = w)

loess_gcomp_reg <- loess(cvd_5year ~ risk_a1, data = test)
loess_gcomp_ipw <- loess(cvd_5year ~ risk_a1, data = test_treated, weights = w)

# Obtain loess predictions for the full test set
test <- test %>%
  mutate(
    cal_steno_reg = predict(loess_steno_reg, newdata = test),
    cal_steno_ipw = predict(loess_steno_ipw, newdata = test),
    cal_gcomp_reg = predict(loess_gcomp_reg, newdata = test),
    cal_gcomp_ipw = predict(loess_gcomp_ipw, newdata = test)
  )

# Reshape data for plotting
cal_steno <- test %>%
  select(pred_steno, cal_steno_reg, cal_steno_ipw) %>%
  pivot_longer(cols = starts_with("cal_"), names_to = "Population", values_to = "observed") %>%
  mutate(
    Population = if_else(Population == "cal_steno_reg", "Observed (Unweighted)", "Counterfactual (IPW-Weighted)"),
    Model = "Predictive Model (Steno)",
    predicted = pred_steno
  ) %>%
  select(predicted, observed, Population, Model)

cal_gcomp <- test %>%
  select(risk_a1, cal_gcomp_reg, cal_gcomp_ipw) %>%
  pivot_longer(cols = starts_with("cal_"), names_to = "Population", values_to = "observed") %>%
  mutate(
    Population = if_else(Population == "cal_gcomp_reg", "Observed (Unweighted)", "Counterfactual (IPW-Weighted)"),
    Model = "Causal Model (G-Formula)",
    predicted = risk_a1
  ) %>%
  select(predicted, observed, Population, Model)

cal_plot_data <- bind_rows(cal_steno, cal_gcomp)

# Plot calibration comparison
ggplot(cal_plot_data, aes(x = predicted, y = observed, color = Population)) +
  geom_line(linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~ Model) +
  labs(
    title = "Counterfactual Calibration Analysis",
    subtitle = "Evaluating calibration in the observed vs. counterfactual (statin-treated) populations",
    x = "Predicted CVD Risk",
    y = "Observed / IPW-Weighted Risk",
    color = "Evaluation Population"
  ) +
  scale_color_manual(values = c("Observed (Unweighted)" = "#D55E00", "Counterfactual (IPW-Weighted)" = "#0072B2")) +
  xlim(0, 0.4) +
  ylim(0, 0.4)
```

#### Calibration Analysis:

1.  **Predictive Model (Steno)**: This model is well-calibrated for the **Observed** population (red line is near the diagonal). However, in the **Counterfactual** population where everyone receives statins (blue line), the model severely overestimates the risk of CVD. This is because it fails to account for the risk-reducing effect of the statin initiation.
2.  **Causal Model (G-Formula)**: This model is poorly calibrated for the **Observed** population (red line deviates) because it predicts risk *under the intervention of statins*, but many patients in the observed cohort were untreated. Crucially, in the **Counterfactual** population (blue line), the model is perfectly calibrated, proving its validity for interventional decision support.

------------------------------------------------------------------------

## 6. Deployment: Causal Projections

We deploy our IPW-weighted causal model (which does not require the polygenic risk score `steno_prs`) on a new patient cohort. We project the risk under both strategies and compute the **Individual Treatment Effect (ITE)**, defined as the absolute risk reduction:

$$\text{ITE}_i = \Pr(Y_i^0 = 1 \mid X_i^*) - \Pr(Y_i^1 = 1 \mid X_i^*)$$

```{r}
#| label: deployment
# Load deployment cohort via public URL
deploy <- readRDS(url("https://smartbiomed.github.io/causal-prediction-course/data/exercises/data/type1-diabetes-deployment.rds"))

# Generate counterfactual predictions using IPW outcome models
deploy <- deploy %>%
  mutate(
    risk_a1 = predict(m_ipw1, newdata = deploy, type = "response"),
    risk_a0 = predict(m_ipw0, newdata = deploy, type = "response"),
    ite = risk_a0 - risk_a1
  )

# Visualize the distribution of individual benefit (ITE)
ggplot(deploy, aes(x = ite)) +
  geom_histogram(binwidth = 0.002, fill = "#0072B2", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  labs(
    title = "Distribution of Individual Treatment Benefit (ITE) in Deployment Cohort",
    subtitle = "ITE = Predicted Risk (No Statin) - Predicted Risk (Statin)",
    x = "Absolute Risk Reduction (Benefit)",
    y = "Number of Patients"
  )
```

```{r}
#| label: deploy-summary
summary(deploy$ite)
```

### Clinical Translation:

The Average Treatment Effect in this cohort is around **1%** risk reduction, but the ITE distribution reveals significant heterogeneity: \* Patients with an ITE close to **0** derive almost no benefit from statins. \* A subset of patients achieves an absolute risk reduction of **2% to 3%** or more. \* A clinical decision rule can be established to recommend statins only if $\text{ITE}_i \ge \text{threshold}$ (e.g., $1\%$), optimizing treatment benefits while minimizing side effects and polypharmacy.

------------------------------------------------------------------------

## Conclusions

1.  **Confounding Bias**: Standard predictive models learn biased associations due to confounding by indication and cannot guide clinical treatment decisions.
2.  **Causal Estimation**: Separating predictors into confounders ($L$) and predictors ($P$) allows the G-formula and IPW to estimate unbiased interventional risks.
3.  **Causal Validation**: Counterfactual Brier scores and weighted calibration curves are essential to validate that causal models perform correctly under the intended treatment scenarios.
4.  **Precision Medicine**: Projecting the Individual Treatment Effect (ITE) enables personalized decision support, targeting treatment to patients who will benefit the most.